Efficient forwards-time simulation with ancestral recombination graphs

### 2017-11-08 forwards_paper.tex
added wf figs
various fig updates

%%%%%%
\subsection*{Tree sequences}
A tree sequence is an efficient encoding for a sequence of correlated trees such as
is produced by recombination. The encoding is efficient because branches that are
shared by adjacent trees are stored once, rather than repeatedly for each tree.
The topology of a tree sequence is defined via a set of \emph{nodes} and
\emph{edges}. A node in a tree sequence refers to a distinct ancestor and
corresponds to the vertices in the individual genealogies along the sequence.
Since each node represents a specific ancestor, it has a unique ``time'',
thought of as her birth time, which determines the height of any vertices
she is associated with. In the example of Figure~\ref{fig:example_tree_sequence}
we have a total of five nodes. Nodes 0, 1 and 2 occur at time 0 and are our
samples. Nodes 3 and 4 represent the ancestors of these samples, and were born at
time 1.5 and 2.5 respectively.

@@

@@

The tree sequence encoding is compact but it is also efficient to process.
Many algorithms to compute statistics of interest for population genetics
are naturally expressed in terms of tree topologies. For example, the
pairwise nucleotide diversity $\pi$, is defined as the average number of
differences between sequences in the sample. Computing $\pi$ directly
from observed sequence data requires $O(n^2 m)$ time, since we must
compare all pairs of samples at every site. However, given the topology of
the trees at every site and the locations of mutations much better algorithms
are possible. Using the fact that we can count the number of samples below
a given node efficiently using previously described tree sequence
algorithms~\citep{kelleher2016efficient}, the time required to compute $\pi$
becomes roughly $O(n \log n + m)$.
The \msprime\ API provides a method to compute $\pi$ among arbitrary subsets of the
samples in a tree sequence, and calculating $\pi$ over all 500K samples
in the example above required about 1.2 seconds.
\jk{I haven't checked the analysis of computing $\pi$ here and not thought about
it too deeply. I guess we should also compare this with something else in order to
show that this time is pretty good. We could get a subset of the sites and push
the data into pylibseq maybe, or scikit-allel??}

@@

\begin{figure}
    \begin{center}
        \includegraphics{wf-before}
        \includegraphics{wf-after}
    \end{center}
    \caption{An example of a marginal genealogy from a Wright-Fisher simulation
    with $N=5$. On the left we have the original tree including all
    intermediate nodes and dead-ends. On the right we have the minimal tree
    relating all of the currently-alive individuals (27--31).
    \jk{This figure needs a little tidying up. We should combine the two SVGs
        into a single image with some (a) (b) annotations. Also the node labels
        are poorly aligned. The SVGs are generated by the sims/wf-figures.py.}
    \label{fig:wf-trees}
    }
\end{figure}

@@

>>>>

%%%%%%
\subsection*{Tree sequences}

A tree sequence is an efficient encoding for a sequence of correlated trees such as
is produced by recombination. The encoding is efficient because branches that are
shared by adjacent trees are stored once, rather than repeatedly for each tree.
The topology of a tree sequence is defined via a set of \emph{nodes} and
\emph{edges}. A node in a tree sequence refers to a distinct ancestor and
corresponds to the vertices in the individual genealogies along the sequence.
Since each node represents a specific ancestor, it has a unique ``time'',
thought of as her birth time, which determines the height of any vertices
she is associated with. In the example of Figure~\ref{fig:example_tree_sequence}
we have a total of five nodes. Nodes 0, 1 and 2 occur at time 0 and are our
samples. Nodes 3 and 4 represent the ancestors of these samples, and were born at
time 1.5 and 2.5 respectively.

@@

Recovering the sequence of trees from a collection of edges is straightforward:
each point along the genome that the tree topology changes
is accompanied by the end of some \emph{edges} and the beginning of others.
Since each \emph{edge} records the genomic interval 
over which a given node inherits from a particular ancestor,
to construct the tree at a certain point in the genome
we need only retrieve all edges overlapping that point
and construct the corresponding edges.
To construct the tree at a nearby location,
we need only remove those edges whose intervals do not overlap that location,
and add those new edges whose intervals do.
Incidentally, this property that edges naturally encode \emph{differences}
between nearby trees (e.g., as ``subtree prune and regraft'' moves)
allows for efficient algorithms that take advantage
of the highly correlated nature of nearby trees.

@@

@@

\begin{figure}
    \begin{center}
        \includegraphics{wf-before-after}
    \end{center}
    \caption{An example of a marginal genealogy from a Wright-Fisher simulation
    with $N=5$. \textbf{(A)} the original tree including all
    intermediate nodes and dead-ends, and \textbf{(B)} the minimal tree
    relating all of the currently-alive individuals (27--31).
    \label{fig:wf-trees}
    }
\end{figure}

@@

This works great and very fast.

Having results in a tree sequence is really good downstream too:
The tree sequence encoding is compact but it is also efficient to process.
Many algorithms to compute statistics of interest for population genetics
are naturally expressed in terms of tree topologies. For example, the
pairwise nucleotide diversity $\pi$, is defined as the average number of
differences between sequences in the sample. Computing $\pi$ directly
from observed sequence data requires $O(n^2 m)$ time, since we must
compare all pairs of samples at every site. However, given the topology of
the trees at every site and the locations of mutations much better algorithms
are possible. Using the fact that we can count the number of samples below
a given node efficiently using previously described tree sequence
algorithms~\citep{kelleher2016efficient}, the time required to compute $\pi$
becomes roughly $O(n \log n + m)$.
The \msprime\ API provides a method to compute $\pi$ among arbitrary subsets of the
samples in a tree sequence, and calculating $\pi$ over all 500K samples
in the example above required about 1.2 seconds.
\jk{I haven't checked the analysis of computing $\pi$ here and not thought about
it too deeply. I guess we should also compare this with something else in order to
show that this time is pretty good. We could get a subset of the sites and push
the data into pylibseq maybe, or scikit-allel??}
